Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-08-27 bsis_paper/paper.tex
* More in Section 2. * Some effort in Section 3.
git-svn-id: file:///home/rblake/svn/pamg/trunk@20 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
Independent set-based coarse grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is shown in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,
                                C,\, \Omega,\, w)$%
end
\end{lstlisting}

@@

Vertex $i$ is eligible to be added to the independent set $D$ if the
following condition is satisfied:
\begin{equation}
  \label{eq:D-selection}
  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
\end{equation}

Many independent set-based coarsening algorithms form $\Omega_{k+1}$
using the following heuristics.
\begin{description}
\item[H1:] For each unknown $j$ that strongly influences $F$-point
  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
  $k$ that also strongly influences $i$.
\item[H2:] The set of $C$-points needs to form a maximal independent
  set in the reduced graph of $S$ such that no $C$-point strongly
  depends on another $C$-point.
\end{description}
H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
enforced by independent set-based methods, while H2 serves as a
guideline whose purpose is to encourage selection of small
$|\Omega_{k+1}|$.

@@

Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

>>>>

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
Many independent set-based coarsening algorithms form $\Omega_{k+1}$
using the following heuristics.
\begin{description}
\item[H1:] For each unknown $j$ that strongly influences $F$-point
  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
  $k$ that also strongly influences $i$.
\item[H2:] The set of $C$-points needs to form a maximal independent
  set in the reduced graph of $S$ such that no $C$-point strongly
  depends on another $C$-point.
\end{description}
H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
enforced by independent set-based methods, while H2 serves as a
guideline whose purpose is to encourage selection of small
$|\Omega_{k+1}|$.

Independent set-based coarse grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is shown in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,
                                C,\, \Omega,\, w)$%
end
\end{lstlisting}

@@

In classical coarsening, the independent set $D$ consists of a single
vertex $i$ in each iteration, where $w_i \ge w_j,\, \forall j \in S_i
\cup S_i^T$. Forming independent sets with only a single vertex is
inefficient in parallel, so parallel independent set methods use
different initialization techniques leading to the ability to form
larger independent sets. In these algorithms vertex $i$ is eligible to
be added to the independent set $D$ if the following condition is
satisfied:
\begin{equation}
  \label{eq:D-selection}
  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
\end{equation}
The research presented here focuses on the independent set criterion
in~(\ref{eq:D-selection}).

The weight update routines used in independent set-based coarsening
algorithms are varied but primarily take one of two forms: routines
that decrease the values of weights and routines that increase the
value of weights. In all cases, a relatively large weight confers
large value for having the corresponding vertex become a $C$-point.

Many independent set-based coarsening algorithms have been
developed. Ruge-St\"uben coarsening, also called classical coarsening,
is introduced in~\cite{Ruge1987}. The first parallel independent
set-based coarsening algorithms appeared approximately ten years ago
and include RS3~\cite{Henson2002}, CLJP~\cite{Cleary1998}, BC-RS and
Falgout~\cite{Henson2002}, PMIS and HMIS~\cite{Sterck2006},
CLJP-c~\cite{alber-cljpc}, PMIS-c1 and PMIS-c2~\cite{alber-PCGS}, and
PMIS Greedy~\cite{Butler2006}, to name a few.

@@

Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{eq:D-selection}), meaning vertex $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood,
which is defined as follows.
\begin{defn}
\label{def:neighborhood}
The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
union of the sets containing the strong influences of $i$ and the
strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
\end{defn}
Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}
